Supplementary Information for ``Engineering spin squeezing in a 3D optical lattice with interacting spin-orbit-coupled fermions''

### 2019-09-13 soc_squeezing/squeezing_paper/main.tex
uploading final changes prior to last submission of squeezing paper

If we cannot take the tunneling rate $J\to0$, our remaining control parameter for turning off SOC is the SOC angle $\phi=k_{\t{L}}a$, where $k_{\t{L}}$ is the projection of the clock laser wavenumber onto the lattice axis and $a$ is the lattice spacing.
The squeezing protocol needs a clock laser with a small but nonzero SOC angle $\phi\ll1$, while the clock interrogation protocol requires a clock laser with $\phi=0$.
Simply using one clock laser with $\phi\ne0$ for the squeezing protocol and another clock laser with $\phi=0$ for the clock interrogation protocol, however, does not resolve this discrepancy, because 
%(i) both the squeezing and clock interrogation protocols need a clock laser that is homogeneous -- i.e.~with Hamiltonian proportional to $S_\x$,
%(i) lasers with different SOC angles are homogeneous in different gauges, and 
%(ii) spin squeezing is a gauge-dependent quantity.
%More precisely,
a state that is squeezed with respect to spin operators that are homogeneous (i.e.~of the form $S_\x,S_\y,S_\z$) in a particular gauge is not necessarily squeezed with respect to spin operators that are homogeneous in a different gauge.
In this appendix, we will work explicitly in the ``lab gauge'' of the clock interrogation protocol, in which the Fermi-Hubbard Hamiltonian is SOC-free and the $\phi=0$ clock laser is homogeneous. To resolve the fact that our squeezing protocol prepares a state that is squeezed in the ``wrong gauge'', we will construct a simple pulse sequence that transforms the inhomogeneous spin operators accessible by the $\phi\ne0$ laser into a homogeneous form in our lab gauge.

>>>>

If we cannot take the tunneling rate $J\to0$, our remaining control parameter for turning off SOC is the SOC angle $\phi=k_{\t{L}}a$, where $k_{\t{L}}$ is the projection of the clock laser wavenumber onto the lattice axis and $a$ is the lattice spacing.
The squeezing protocol needs a clock laser with a small but nonzero SOC angle $\phi\ll1$, while the clock interrogation protocol requires a clock laser with $\phi=0$.
Simply using one clock laser with $\phi\ne0$ for the squeezing protocol and another clock laser with $\phi=0$ for the clock interrogation protocol, however, does not resolve this discrepancy, because 
%(i) both the squeezing and clock interrogation protocols need a clock laser that is homogeneous -- i.e.~with Hamiltonian proportional to $S_\x$,
%(i) lasers with different SOC angles are homogeneous in different gauges, and 
%(ii) spin squeezing is a gauge-dependent quantity.
%More precisely,
a state that is squeezed with respect to spin operators that are homogeneous (i.e.~of the form $S_\x,S_\y,S_\z$) in a particular gauge is not necessarily squeezed with respect to spin operators that are homogeneous in a different gauge.
In this analysis, we will work explicitly in the ``lab gauge'' of the clock interrogation protocol, in which the Fermi-Hubbard Hamiltonian is SOC-free and the $\phi=0$ clock laser is homogeneous. To resolve the fact that our squeezing protocol prepares a state that is squeezed in the ``wrong gauge'', we will construct a simple pulse sequence that transforms the inhomogeneous spin operators accessible by the $\phi\ne0$ laser into a homogeneous form in our lab gauge.

### 2019-09-13 soc_squeezing/squeezing_paper/supplementary.tex
uploading final changes prior to last submission of squeezing paper

%%% physics and math packages
\usepackage{physics} % general physics package
\usepackage{amssymb} % math fonts and symbols
\usepackage{bm} % for making math symbols bold
\usepackage{braket} % for nice brackets (e.g. |X>)
\usepackage{mathtools} % for \coloneqq
\usepackage{graphicx} % for figures
\graphicspath{{./figures/}} % set path for all figures

@@

Current 3D optical lattice implementations involve a harmonic confining potential, which significantly alters the underlying single-particle eigenstructure and can potentially degrade accessible squeezing within our protocol. In this appendix, we examine the effect of a harmonic trap on our protocol and discuss strategies to mitigate undesired effects. We model the trap by the addition of the term
\begin{align}
  \hat{H}_{\Omega}
  = \Omega\sum_{j,\alpha}(j-j_0)^2\hat{n}_{j,\alpha}
\end{align}
to our Fermi-Hubbard model, where $j_0$ denotes the trap center and $\Omega = m(\omega a)^2/2$ characterizes the trap strength for atom mass $m$, trap frequency $\omega$, and lattice spacing $a$. In current state-of-the-art 3D $^{87}$Sr OLC implementations, values of $\omega \approx 56 \times 2\pi~\t{sec}^{-1}$ can be achieved within each 2D layer of weak SOC by utilizing in-plane lattice depths of $5 E_R$ and a lattice depth of $60 E_R$ in the axial direction, resulting in a value of $\Omega/J \approx 0.01$. We restrict our discussion to 1D, although for a separable 3D lattice our arguments should extend in a straightforward manner.

@@

Inelastic $^{3}\text{P}_{0}$ (electronic state $e$ or $\up$) collisions are detrimental for optical lattice clocks. For the  nuclear-spin-polarized gas discussed in this work, $ee$ losses are only possible via the $p$-wave scattering channel since $s$-wave collisions are suppressed by Fermi statistics. The big advantage here compared to  prior experiments done in a 1D lattice at $\mu$K temperature\cite{martin2013quantum} is that in a Fermi  degenerate  gas loaded in a 3D optical lattice, $p$-wave losses are further suppressed by the centrifugal barrier and Pauli blocking, and only happen through a wave-function overlap between atoms at different lattice sites.  In this appendix, we quantify the effect of $p$-wave interactions on squeezing. To account for $p$-wave losses, we describe the dynamics  using a master equation for the system's density matrix $\hat{\rho}$:
\begin{align}
  \frac{d\hat{\rho}}{dt}
  = -i [\hat{H}_{\text{eff}},\,\hat{\rho}]
  + \mathcal{L}\hat{\rho},
\end{align}
where $\hat{H}_{\text{eff}}=\chi\hat{S}_z^2$ is the effective one-axis twisting Hamiltonian obtained from  the original Fermi-Hubbard Hamiltonian with spin-orbit coupling, and $\mathcal{L}$ is the Lindblad superoperator that accounts for $p$-wave $ee$ inelastic collisions. This latter term can be written using a pseudo-potential approximation as\cite{rey2014probing}:
\begin{align}
  \mathcal{L}\hat\rho
  = \sum_{{\bm k}{\bm k'}} \Gamma_{{\bm k}{\bm k'}}
  \sp{\hat{A}_{{\bm k}{\bm k'}} \hat\rho\hat{A}_{{\bm k}{\bm k'}}^\dag
  - \f12 \p{\hat{A}_{{\bm k}{\bm k'}}^\dag \hat{A}_{{\bm k}{\bm k'}} \hat\rho
+\hat\rho\hat{A}_{{\bm k}{\bm k'}}^\dag \hat{A}_{{\bm k}{\bm k'}}}},
\end{align}
where the jump operators are $\hat{A}_{{\bm k}{\bm k'}}=\hat{c}_{{\bm k},\up}\hat{c}_{{\bm k'},\up}$, and
$\bm k$, $\bm k'$ sum over all the populated quasi-momentum modes. The decay matrix elements $\Gamma_{{\bm k}{\bm k'}}$ are given by:
\begin{align}
  \Gamma_{{\bm k}{\bm k'}}
  = \frac{3\pi b_{\text{im}}^3 }{m} \p{\int d{\bm r}^{\,3}
    W[\phi_{{\bm k}}({\bm r}),\phi_{{\bm k'}}({\bm r})]},
\end{align}
where $b_{\text{im}} = 121 a_0$\cite{zhang2014spectroscopic, goban2018emergence} is the $p$-wave inelastic scattering length (with $ a_0=5.29\times 10 ^{-11}$ m the Bohr radius), $\phi_{{\bm k}}({\bm r})$ is the Bloch function with quasi-momentum ${\bm k}$, and
\begin{multline}
  W\sp{\phi_{{\bm k}}({\bm r}),\phi_{{\bm k'}}({\bm r})} \\
  \equiv \sp{\p{{\bm \nabla}\phi^{*}_{{\bm k}}(\bm r)}
  \phi^{*}_{{\bm k'}}(\bm r)-\phi^{*}_{{\bm k}}(\bm r)
  \p{{\bm \nabla}\phi^{*}_{{\bm k'}}(\bm r)}}
  \cdot\sp{\p{{\bm \nabla}\phi_{{\bm k}}(\bm r)}
  \phi_{{\bm k'}}(\bm r)-\phi_{{\bm k}}(\bm r)
  \p{{\bm \nabla}\phi_{{\bm k'}}(\bm r)}}
\end{multline}

>>>>

%%% physics and math packages
\usepackage{physics} % general physics package
\usepackage{amssymb} % math fonts and symbols
\usepackage{bm} % for making math symbols bold
\usepackage{braket} % for nice brackets (e.g. |X>)
\usepackage{mathtools} % for \coloneqq
\usepackage{graphicx} % for figures
\graphicspath{{../figures/}} % set path for all figures

@@

Current 3D optical lattice implementations involve a harmonic confining potential, which significantly alters the underlying single-particle eigenstructure and can potentially degrade accessible squeezing within our protocol. In this section of the supplement, we examine the effect of a harmonic trap on our protocol and discuss strategies to mitigate undesired effects. We model the trap by the addition of the term
\begin{align}
  \hat{H}_{\Omega}
  = \Omega\sum_{j,\alpha}(j-j_0)^2\hat{n}_{j,\alpha}
\end{align}
to our Fermi-Hubbard model, where $j_0$ denotes the trap center and $\Omega = m(\omega a)^2/2$ characterizes the trap strength for atom mass $m$, trap frequency $\omega$, and lattice spacing $a$. In current state-of-the-art 3D $^{87}$Sr OLC implementations, values of $\omega \approx 56 \times 2\pi~\t{sec}^{-1}$ can be achieved within each 2D layer of weak SOC by utilizing in-plane lattice depths of $5 E_R$ and a lattice depth of $60 E_R$ in the axial direction, resulting in a value of $\Omega/J \approx 0.01$. We restrict our discussion to 1D, although for a separable 3D lattice our arguments should extend in a straightforward manner.

@@

Inelastic $^{3}\text{P}_{0}$ (electronic state $e$ or $\up$) collisions are detrimental for optical lattice clocks. For the  nuclear-spin-polarized gas discussed in this work, $ee$ losses are only possible via the $p$-wave scattering channel since $s$-wave collisions are suppressed by Fermi statistics. The big advantage here compared to  prior experiments done in a 1D lattice at $\mu$K temperature\cite{martin2013quantum} is that in a Fermi  degenerate  gas loaded in a 3D optical lattice, $p$-wave losses are further suppressed by the centrifugal barrier and Pauli blocking, and only happen through a wave-function overlap between atoms at different lattice sites.  In this supplementary material, we quantify the effect of $p$-wave interactions on squeezing. To account for $p$-wave losses, we describe the dynamics  using a master equation for the system's density matrix $\hat{\rho}$:
\begin{align}
  \frac{d\hat{\rho}}{dt}
  = -i [\hat{H}_{\text{eff}},\,\hat{\rho}]
  + \mathcal{L}\hat{\rho},
\end{align}
where $\hat{H}_{\text{eff}}=\chi\hat{S}_z^2$ is the effective one-axis twisting Hamiltonian obtained from  the original Fermi-Hubbard Hamiltonian with spin-orbit coupling, and $\mathcal{L}$ is the Lindblad superoperator that accounts for $p$-wave $ee$ inelastic collisions. This latter term can be written using a pseudo-potential approximation as\cite{rey2014probing}:
\begin{align}
  \mathcal{L}\hat\rho
  = \sum_{{\bm k}{\bm k'}} \Gamma_{{\bm k}{\bm k'}}
  \sp{\hat{A}_{{\bm k}{\bm k'}} \hat\rho\hat{A}_{{\bm k}{\bm k'}}^\dag
  - \f12 \p{\hat{A}_{{\bm k}{\bm k'}}^\dag \hat{A}_{{\bm k}{\bm k'}} \hat\rho
+\hat\rho\hat{A}_{{\bm k}{\bm k'}}^\dag \hat{A}_{{\bm k}{\bm k'}}}},
\end{align}
where the jump operators are $\hat{A}_{{\bm k}{\bm k'}}=\hat{c}_{{\bm k},\up}\hat{c}_{{\bm k'},\up}$, and
$\bm k$, $\bm k'$ sum over all the populated quasi-momentum modes. The decay matrix elements $\Gamma_{{\bm k}{\bm k'}}$ are given by:
\begin{align}
  \Gamma_{{\bm k}{\bm k'}}
  = \frac{3\pi b_{\text{im}}^3 }{m} \p{\int d{\bm r}^{\,3}
    W[\phi_{{\bm k}}({\bm r}),\phi_{{\bm k'}}({\bm r})]},
\end{align}
where $b_{\text{im}} = 121 a_0$\cite{zhang2014spectroscopic, goban2018emergence} is the $p$-wave inelastic scattering length (with $ a_0=5.29\times 10 ^{-11}$ m the Bohr radius), $\phi_{{\bm k}}({\bm r})$ is the Bloch function with quasi-momentum ${\bm k}$, and
\begin{multline}
  W\sp{\phi_{{\bm k}}({\bm r}),\phi_{{\bm k'}}({\bm r})} \\
  \equiv \sp{\p{{\bm \nabla}\phi^{*}_{{\bm k}}(\bm r)}
  \phi^{*}_{{\bm k'}}(\bm r)-\phi^{*}_{{\bm k}}(\bm r)
  \p{{\bm \nabla}\phi^{*}_{{\bm k'}}(\bm r)}}
  \cdot\sp{\p{{\bm \nabla}\phi_{{\bm k}}(\bm r)}
  \phi_{{\bm k'}}(\bm r)-\phi_{{\bm k}}(\bm r)
  \p{{\bm \nabla}\phi_{{\bm k'}}(\bm r)}}
\end{multline}
